% Contact Thermodynamics — Conference Paper Draft
% January 18, 2026
\documentclass[10pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}

\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{teal},
  breaklines=true,
  frame=single,
  columns=flexible
}

\title{Contact Thermodynamics: A JavaScript Library for Contact Hamiltonian Dynamics on 1-Jet Bundles}
\author{Ginanjar Utama \\ Indonesia \\ \texttt{ginanjar.utama@gmail.com}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We present \emph{Contact Thermodynamics}, an open-source JavaScript library that implements contact geometry on 1-jet bundles for extended thermodynamics, wave mechanics, and contact Hamiltonian dynamics. The library provides canonical contact manifolds, contact Hamiltonian vector fields, Runge--Kutta integration, Legendrian submanifolds, and a relativistic extension through curved spacetime Hamiltonians. The latest version extends this foundation with \emph{Geometric Algebra} (Clifford Algebra) and discrete geometric calculus, enabling field theoretic simulations on contact manifolds. This draft describes the mathematical formulation, software architecture, and representative usage including 3D electromagnetic wave simulation.
\end{abstract}

\section{Introduction}
Contact geometry provides the natural mathematical framework for systems with dissipation, thermodynamic potentials, and odd-dimensional phase spaces. Unlike symplectic mechanics, where the Hamiltonian is conserved, contact Hamiltonian dynamics permits non-conservation consistent with irreversible processes. This paper introduces \emph{Contact Thermodynamics}, a JavaScript implementation targeting rapid experimentation, teaching, and interactive visualization. 

\paragraph{Contributions.} The library provides: (i) a full 1-jet bundle model of contact manifolds, (ii) contact Hamiltonian dynamics with numerical integration, (iii) a complete Geometric Algebra module for Cl($p,q$), (iv) a discrete differential calculus engine for solving wave equations, and (v) interactive browser demos including 3D electromagnetic wave simulation.

\section{Mathematical Background}
\subsection{Contact Geometry}
Let $Q$ be a configuration manifold of dimension $n$. The 1-jet bundle $J^1(Q)$ has canonical coordinates $(x^a, u, p_a)$ with dimension $2n+1$. The canonical contact form is
\begin{equation}
\alpha = du - p_a\,dx^a.
\end{equation}
A contact manifold satisfies the non-degeneracy condition $\alpha\wedge(d\alpha)^n\neq 0$.

Given a Hamiltonian $H: M\to\mathbb{R}$, the contact Hamiltonian vector field $X_H$ satisfies
\begin{align}
\iota_{X_H}\alpha &= -H, \\
\iota_{X_H} d\alpha &= dH - (R H)\,\alpha.
\end{align}
In canonical coordinates, this yields the dissipative equations of motion:
\begin{align}
\dot{x}^a &= \frac{\partial H}{\partial p_a}, \quad
\dot{p}_a = -\frac{\partial H}{\partial x^a} - p_a\,\frac{\partial H}{\partial u}, \quad
\dot{u} = p_a\frac{\partial H}{\partial p_a} - H.
\end{align}

\subsection{Geometric Algebra}
The library now incorporates Clifford Algebra $\text{Cl}(p,q)$ to handle geometric primitives and transformations. Elements are multivectors $A \in \Lambda(V)$, and the product is defined as $ab = a \cdot b + a \wedge b$. Rotations and boosts are generated by rotors $R = \exp(-B\phi/2)$, where $B$ is a bivector. We classify bivectors $B$ by their square $B^2$:
\begin{itemize}
    \item $B^2 < 0$: Elliptic (rotations) $\to \cos, \sin$
    \item $B^2 > 0$: Hyperbolic (boosts) $\to \cosh, \sinh$
    \item $B^2 = 0$: Parabolic (translations)
\end{itemize}

\section{Library Design and API}
The library is modular, separating the contact geometry manifold structure from the geometric algebra compute engine and discrete calculus solvers.

\subsection{Contact Manifolds}
Three standard models are included:
\begin{itemize}
    \item \textbf{Grand Model} $M_{13}$: full phase space with six base coordinates.
    \item \textbf{Holographic Model} $M_7$: reduced base $(t,\ell,S)$ with emergent spatial fields.
    \item \textbf{Gauge-Extended Model} $M_{15}$: adds gauge phase $(\varphi,I)$.
\end{itemize}

\subsection{Differential Geometry \& Field Theory}
New modules enable covariant field theory on these manifolds:
\begin{itemize}
    \item \texttt{ChristoffelSymbols}: Computes $\Gamma^k_{ij}$ from any metric $g_{\mu\nu}$.
    \item \texttt{CovariantDerivative}: Implements $\nabla_i$ and parallel transport.
    \item \texttt{SplitDifferentialOperator}: Implements the discrete operator $\nabla = \sum e^i \partial_i$ for mesh-based simulations.
\end{itemize}

\begin{lstlisting}[language=JavaScript,caption={Maxwell solver using discrete geometric calculus.}]
// Faraday: dB/dt = -curl(E)
const dEz_dy = (Ez.get(i, j+1) - Ez.get(i, j-1)) / (2*dx);
Bx.set(Bx.get(i,j) - dt * dEz_dy, i, j);

// Ampere: dE/dt = c^2 curl(B)
const curlB = calculateCurl(Bx, By);
Ez.set(Ez.get(i,j) + c2dt * curlB, i, j);
\end{lstlisting}

\section{Visualization and Demos}
The project includes interactive browser demos. The newest addition is a \textbf{3D Electromagnetic Wave Simulation} using the discrete geometric calculus module. It solves Maxwell's equations on a grid and renders the $E$-field envelope in real-time ($>60$ FPS) using WebGL, demonstrating energy conservation in the contact-geometric setting.

\section{Relativistic Extension}
The library supports relativistic Hamiltonians coupling contact structure to curved spacetime metrics.
\begin{equation}
H = \frac{1}{2}g^{\mu\nu}(x)(p_\mu - qA_\mu)(p_\nu - qA_\nu) - \frac{1}{2}m^2 = 0.
\end{equation}
Parallel transport verification tests confirm the path-dependence (holonomy) of vector transport on curved manifolds like the 2-sphere and Schwarzschild spacetime.

\section{Conclusion}
Contact Thermodynamics lowers the barrier to using contact geometry and geometric algebra in computational experiments. By integrating strict jet-bundle geometry with versatile Clifford algebra operators, it enables the simulation of complex dissipative and field-theoretic systems directly in the browser.

\section*{Availability}
Source code and documentation are available at \url{https://github.com/gutama/contact-thermodynamics}.

\begin{thebibliography}{9}
\bibitem{geiges2008} H. Geiges, \emph{An Introduction to Contact Topology}. Cambridge, 2008.
\bibitem{bravetti2017} A. Bravetti, ``Contact Hamiltonian Dynamics,'' \emph{Entropy}, 2017.
\bibitem{hestenes2002} D. Hestenes, \emph{New Foundations for Classical Mechanics}, Kluwer, 2002.
\bibitem{doran2003} C. Doran \& A. Lasenby, \emph{Geometric Algebra for Physicists}, Cambridge, 2003.
\end{thebibliography}

\end{document}
